`flippant` -- An `R` package for the automated  analysis of dithionite

### 2015-09-16 inst/manuscript/manuscript.rmd
Finalize reference addition to first intro draft and keep paraphrasing editing; This is the first version sent to Birgit for feedback;

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
lipids between the leaflets of lipid bilayers representing biological membranes
[@waechter_flippases_2013]. Indipendent of energy equivalents such as ATP they 
facilitate this ``flipping'' by providing a mechanism of transition through the 
hydrophobic environment of membranes to polar lipid head groups, thus 
reducing the massive energetic cost that process would carry otherwise <CITE 
ANANT HALF TIME>. While the underlying molecular mechanisms remain enigmatic 
<CITE SOMETHING>, and molecules supporting flipping processes as central as the 
equilibration of the ubiquitous phosphoglycerolipids across the membrane of the 
endoplasmic reticulum, where they are synthesized in the cytoplasmic leaflet, 
have yet to be identified <CITE SOMETHING>, there has been much recent progress 
in scramblase identification and characterization. Molecules shown to have 
scramblase activity include opsin/rhodopsin 
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. **Would be good to include/replace some 
with non-in-house identifications/characterizations. What else is out there?**

To interrogate factors of this functional class, we have developed a biochemical 
assay suited to characterize scramblase activity in purified proteins and protein 
extracts [@chang_chemical_2004], based on the reconstitution of scramblases or 
candidate proteins into large synthetic vesicles with a lipid
composition that includes nitrobenzoxadiazole (NBD) labeled fluorescent lipids.
Upon exposure of these vesicles to membrane-impermeable reducing agents such as 
dithionite, fluorophores in the external leaflet of the vesicles are quenched and 
fluorescence in scramblase-free vesicles accordingly quickly drops to 
approximately 50% of the initial value. However, for vesicles reconstituted with
proteins conferring a scramblase activity, fluorescence dissapears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the vesicle membrane renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the vesicle membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@goren_constitutive_2014;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014],
the assay has for example 
also been used to characterize the stoichiometry of scramblase homo-multimers 
reconstituting into the vesicles [@ploier_perturbed_submitted] and probe the 
ATP-dependence of flippases, molecules that under energy-consumption flip lipids
against a concentration gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the package ...**

Starting with the identification of baseline and post-reduction fluorescense from 
fluorometer-generated spectra, via the calculation of the underlying statistics
and through publication-grade representation of the results, data analysis in 
the context of this assay has until now involved an extensive series of manual 
steps, at most supported by spreadsheet calculation facilities [see the 
supplementary material of @menon_opsin_2011 for extensive stepwise discussion]. Based on the
`R` free software environment for statistical computing and graphics [@r_core_team_r:_2015]
and prominently drawing on the power and elegance of the `ggplot2` plotting 
system [@wickham_ggplot2_2009], we here provide `flippant`, a comprehensive package
covering fast, reproducible data analysis and graphing for scramblase assays 
from raw fluorescense spectra with a simple parametrizing spreadsheet as the
sole required user input.

>>>>

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
lipids between the leaflets of lipid bilayers that constitute biological membranes
[@waechter_flippases_2013]. Indipendent of energy equivalents such as ATP they 
facilitate this ``flipping'' by providing to polar lipid head groups a mechanism
of transition through the hydrophobic environment of membranes, thus reducing 
the massive energetic cost that process would carry otherwise 
[@kornberg_inside-outside_1971]. While the underlying molecular mechanisms in 
many cases remain enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015], and molecules supporting flipping processes as 
central as the equilibration of the ubiquitous phosphoglycerolipids across the 
membrane of the endoplasmic reticulum, where they are synthesized in the 
cytoplasmic monolayer, have yet to be identified **(I was looking for a recent 
citation that clearly spelled that out - in vain. Do you have a proposal?)**, 
there has been much recent progress in scramblase identification and 
characterization. Molecules shown to have scramblase activity include 
opsin/rhodopsin [@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. **Would be good to include/replace some 
with non-in-house identifications/characterizations. What else is out there?**

To interrogate factors of this functional class, we have developed a biochemical 
assay suited to characterize scramblase activity in purified proteins and protein 
extracts [@chang_chemical_2004] based on the reconstitution of scramblases or 
candidate proteins into synthetic liposomes with a lipid
composition that includes nitrobenzoxadiazole (NBD) labeled fluorescent lipids.
Upon exposure of these liposomes to membrane-impermeable reducing agents such as 
dithionite, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence dissapears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@goren_constitutive_2014;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the assay has also been used to characterize the stoichiometry of scramblase homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the 
package ...**

Starting with the identification of baseline and post-reduction fluorescense 
from fluorometer-generated spectra, via the calculation of the underlying 
statistics and through publication-grade representation of the results, data 
analysis in the context of this assay has until now involved an extensive series
of manual steps, at most supported by spreadsheet calculation facilities
[see the supplementary material of @menon_opsin_2011 for extensive stepwise discussion].
Based on the `R` free software environment for statistical computing and 
graphics [@r_core_team_r:_2015] and prominently drawing on the power and 
elegance of the `ggplot2` plotting system [@wickham_ggplot2_2009], we here 
provide `flippant`, a package that comprehensively, fast and reproducibly covers
all data analysis and graphing needs from raw fluorescense spectra produced by 
scramblase assays. A simple parametrizing spreadsheet is the sole required user 
input.
